The Evolution of Phenotypically Invariant Genetic Networks

### 2017-06-20 handnotes.tex
more words about variation

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know two things:
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  Both are hard to measure, somewhat ambiguous, and have values that may depend strongly on the context.
  We aim to get order-of-magnitude estimates.

  Specifically, we want to know how much the entries of $A$ vary within a population.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  Furthermore, the mutational target size for variation in $A$ may be much larger than the kilobase or so of a typical promoter sequence.
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}

  The amount and structure of this standing variation is established over long time scales
  by many factors, including
  mutation-selection balance, 
  shifts in the phenotypic optimum,
  and/or spatial variation in the optimum \citep{hansen1996translating}.
  Quantitative genetics models of mutation-selection balance 
  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
  but it is unclear how well these predictions match reality \citep{johnson_barton}
  and how much they are expected to change over time \citep{arnold_changing_G}.
  However, empirical work allows us to estimate at least the rough magnitude of variation.
  XXX FILL IN DETAILS: 
  Mutational scans find this distribution of binding site change per nucleotide change in promoters.
  Intraspecific variation is often in the range of 0.1\%--1\%.
  This suggests typical regulatory binding strengths within a population might vary by up to roughly X\%.
  But, does it?
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila,
  which suggests that (a) the observed range of variation is somewhat less than this,
  and (b) the strength of selection on phenotype is sufficient to weakly constrain this variation.

  This suggests that within a population, transcription factor binding strengths 
  -- the entries of $A$ --
  vary by around X\%, at least for networks whose function is strongly constrained.
  Subsequent generations for the most part resample from this diversity,
  so in a population of effective size $N_e$,
  simply by the variance of the mean of a random sample,
  the population mean binding strengths
  will move a few multiples of $X/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
  (This could be taken as a definition of $N_e$.)
  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
  unless recombination load is substantial \citep{recomb_load}.
  The action of genetic drift is also strongly determined by covariance between 
  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
  covariance which may arise due to functional constraints and/or statistical linkage.
  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
  Linkage will almost certainly lead to covariance
  if the variation is due to \textit{cis}-regulatory variants,
  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
  across which a population may carry only a few common haplotypes.
  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
  In principle, there may be substantially less variation away from the set of optimal coefficients 
  than there is along the set, due to the action of selection.
  If so, this might require substantial epistatic load 
  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
  However, it is unclear if this is likely to occur and XXX we revisit this below.

@@

  Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation. \citet{tulchinsky } simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybird incomaptibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproducitve isolation primarily following direcitonal selection for a change in expression levels in each allopatric population; the evidecne for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overal network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, shapes}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)
\usepackage{authblk}

@@

  At any given time, there will be a range of network coefficients present in the population
  due to segregating genetic polymorphism.
  Over many generations, even if selective pressures do not change,
  this range of networks will shift 
  as recombination, mutation, and demographic noise create new alleles and shift allele frequencies.
  How much variation do we expect to find within a population?
  Is this range limited by available variation or kept in check by selection?
  How fast will a population explore the space of equivalent networks?
  To answer these questions, we need to know two things known at best only roughly \citep{felsenstein1988phylogenies} --
  the strength of stabilizing selection on the phenotype,
  and the amount (and structure) of heritable variation in the genotype.
  We aim to get order-of-magnitude estimates.

  \plr{amount of heritable variation}
  We quantify (roughly) the amount of heritable variation
  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
  For this reason, 
  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}
  Differences in $A_{ij}$ due to a sequence change are hard to measure 
  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
  or in expression levels (e.g., cis-eQTL).
  The former may be an overestimate, 
  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
  \citet{kasowski2010variation} found differential occupancy in 7.5\%
  of binding sites of a transcription factor (p65) between human individuals.
  \citet{verlaan2009targeted} showed that cis-regulatory variation
  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
  while \citep{lappalainen2013transcriptome} found that human population variation 
  explained about 3\% of expression variation.
  \plr{Get some data from at least one other species in here!}
  Taken together, this suggests that variation in the entries of $A$
  may be on the scale of 1\% between individuals of a population --
  doubtless varying substantially between species and between genes.

  \plr{merge this above}
  The amount and structure of this standing variation is established over long time scales
  by many factors, including
  mutation-selection balance, 
  shifts in the phenotypic optimum,
  and/or spatial variation in the optimum \citep{hansen1996translating}.
  Quantitative genetics models of mutation-selection balance 
  predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
  but it is unclear how well these predictions match reality \citep{johnson_barton}
  and how much they are expected to change over time \citep{arnold_changing_G}.
  However, empirical work allows us to estimate at least the rough magnitude of variation.

  \plr{strength of selection}
  It seems certain that selection is not so strong that intrapopulation variation
  is strongly deleterious,
  so that if $u$ is the typical scale on which selection acts
  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
  then $u > \sigma$.
  However,
  a range of studies \plr{find them} have found evidence for weak stabilizing selection
  on regulatory SNPs and cis-eQTL.
  For instance, 
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
  This suggests that 
  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
  so that perhaps $\sigma$ and $u$ are relatively close.
  This is as would be expected if available variation is held in check by mutation--selection balance
  rather than genetic drift.
  A conservative estimate would be that $u = 5 \sigma$;
  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
  to effect a noticable drop in fitness.
  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}

  We have guessed that within a population the entries of $A$
  vary by a factor of $\sigma=0.01$, at least for networks whose function is strongly constrained.
  Subsequent generations for the most part resample from this diversity,
  so in a population of effective size $N_e$,
  simply by the variance of the mean of a random sample,
  the population mean binding strengths
  will move a few multiples of $\sigma/\sqrt{N_e}\%$ per generation \citep{lande_drift}.
  (This could be taken as a definition of $N_e$.)
  Selection will tend to push this mean towards the optimal set of networks \citep{ou_process},
  but mean movement parallel to the optimal set (that leaves the phenotype invariant) is unconstrained
  unless recombination load is substantial \citep{recomb_load}.
  The action of genetic drift is also strongly determined by covariance between 
  standing genetic variation in different regulatory coefficients -- known as the $G$ matrix \citep{G_matrix},
  covariance which may arise due to functional constraints and/or statistical linkage.
  There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
  Linkage will almost certainly lead to covariance
  if the variation is due to \textit{cis}-regulatory variants,
  in which case the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
  across which a population may carry only a few common haplotypes.
  However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
  a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
  (leading to positive covariance between $A_{ik}$ and $A_{jk}$)
  is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
  (Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)
  In principle, there may be substantially less variation away from the set of optimal coefficients 
  than there is along the set, due to the action of selection.
  If so, this might require substantial epistatic load 
  -- mortality or reduced fitness of a large proportion of new offspring due to recombination between somewhat incompatible alleles.
  However, it is unclear if this is likely to occur and XXX we revisit this below.

@@

  Over the last several years, several different computational approaches have been applied to study reproductive incompatibility and speciation. \citet{tulchinsky} simulated the evolution of a transcription factor and its binding site using a thermodynamic model. Their simulations suggest that the language by which a transcription factor recognizes its binding site can change, and potentially lead to hybird incomaptibility when allopatric populations employ divergent readout languages. This study, despite looking at gene regulation, does not analyze overall gene network architecture -- as we do here -- it only looks at the expression level of a single gene. Furthermore, they report reproducitve isolation primarily following direcitonal selection for a change in expression levels in each allopatric population; the evidecne for reproductive isolation following balancing selection is much weaker. Johnson and Porter 2000 did not observe any hybrid fitness declines under stabilizing selection -- only under directional selection. 
  Khatari et al, Tulchinsky et al, and Porter et al, all study hybrid incompatibility from a transcription factor/binding site interaction perspective, not from an overal network architecture perspective. Palmer and Feldman only see hybrid incompatibility in constant environments if the parental populations are relatively poorly adapted initially. Otherwise hybrids between two allopatric populations have fairly high fitnesses.

### 2017-06-20 promoterdata.tex
more words about variation

\documentclass[9 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)

@@

@@

    \item \citet{schmidt2010vertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
        and find (a) conservation of binding motif; (b) little sharing of actual binding sites; and (c) a fair amount of compensatory changes
        \begin{itemize}

@@

    \item \citet{tugrul2015dyanmics} do theory on the gain and loss of TFBS.  
        Conclusion: more than 10bp unlikely; de novo very hard but made possible ad the scale of eukaryotic species divergence by pre-existing relics, cooperativity, etc.
        Euk TFBS typically 6--12bp.

@@

  \item \citet{rebeiz2009stepwise} look at adaptive melaninism in D. melanogaster. 
    Find a minimum of 5 mutations differentiate regulation of dark from light lines. These account for 8 to 40\% of the overal difference in ebony activity.

>>>>

\documentclass{article}
\usepackage{amsmath, amssymb, color, xcolor, amsthm}
\usepackage{graphicx, wrapfig, float, caption, dsfont, bbm}
\usepackage{fullpage}
\usepackage[backref=page, hidelinks, colorlinks=true, citecolor=blue!60!black!100]{hyperref}
\usepackage{caption, subcaption}
\usepackage{natbib} % gives us \citet: Author (year) and \citep: (Author; year)

@@

Generally, there are three sorts of data on variation:
\begin{enumerate}
    \item how binding site occupancy varies in a population
    \item how promoter SNPs affect expression (or phenotype) by swapping them out
    \item how cis-eQTL affect expression by population scans, esp in heterozygotes
\end{enumerate}

\emph{Orphaned, unconvincing paragraph:}
  In \textit{Drosophila}, \citet{frankel2011morphological} 
  found that clusters of 2--5 mutations in enhancers
  reduced expression \plr{of what?} by 4.6--33.5\%,
  and \citet{rebeiz2009stepwise} found that around five cis-regulatory SNPs
  cause differences of 8--40\% of \textit{ebony} expression differences
  between light and dark lines;
  while in human (ENCODE) data, \citet{shi2016evaluating} found that shifts in binding site motif score
  of 5\%--10\% are responsible for differential occupancy.
  These latter observations suggest that changes of several base pairs
  can shift occupancy and/or expression by tens of percent;
  if a typical promoter contains 100 bases relevant for binding of a particular transcription factor,
  in which region 1\% of the population contains a SNP,
  this suggests population expression variance due to such factors of around 0.1--1\%.

@@

    \item \citet{schmidt2010fivevertebrate} use ChIP-seq to compare TF occupancy (for CEBPA and HNF4A) between five vertebrates (human, mouse, dog, opossum, and chicken), 
        and find (a) conservation of binding motif; (b) little sharing of actual binding sites; and (c) a fair amount of compensatory changes
        \begin{itemize}

@@

    \item \citet{tugrul2015dynamics} do theory on the gain and loss of TFBS.  
        Conclusion: more than 10bp unlikely; de novo very hard but made possible ad the scale of eukaryotic species divergence by pre-existing relics, cooperativity, etc.
        Euk TFBS typically 6--12bp.

@@

  \item \citet{rebeiz2009stepwise} look at adaptive melaninism in D. melanogaster. 
      Find a minimum of 5 mutations differentiate regulation of dark from light lines. These account for 8 to 40\% of the overal difference in ebony activity.  \emph{(But, this is adaptive -- different selection in different places.)}
